Libraries
```{r}
library(data.table)
library(psych) 
library(GPArotation)
library(corpcor)
library(EFAtools)
library(lavaan)
library(lavaanPlot)
library(esemComp)
library(ggplot2)
```
Set working directory
```{r}
setwd('/Cog-Ment/R/g_factor_5_folds/')
```
#Upload train and test sets generated and scaled in Python
```{r}
#0
train_0 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_0/train_scaled_fold_0.csv", header=TRUE)
test_0 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_0/test_scaled_fold_0.csv", header=TRUE)
#1
train_1 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_1/train_scaled_fold_1.csv", header=TRUE)
test_1 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_1/test_scaled_fold_1.csv", header=TRUE)
#2
train_2 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_2/train_scaled_fold_2.csv", header=TRUE)
test_2 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_2/test_scaled_fold_2.csv", header=TRUE)
#3
train_3 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_3/train_scaled_fold_3.csv", header=TRUE)
test_3 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_3/test_scaled_fold_3.csv", header=TRUE)
#4
train_4 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_4/train_scaled_fold_4.csv", header=TRUE)
test_4 = read.csv("/Cog-Ment/g_factor_5_folds_python/fold_4/test_scaled_fold_4.csv", header=TRUE)
```
#Factorability
##Bartlett test
```{r}
# Train
cortest.bartlett(cor(train_0), n = nrow(data))
cortest.bartlett(cor(train_1), n = nrow(data))
cortest.bartlett(cor(train_2), n = nrow(data))
cortest.bartlett(cor(train_3), n = nrow(data))
cortest.bartlett(cor(train_4), n = nrow(data))
#Test
cortest.bartlett(cor(test_0), n = nrow(data))
cortest.bartlett(cor(test_1), n = nrow(data))
cortest.bartlett(cor(test_2), n = nrow(data))
cortest.bartlett(cor(test_3), n = nrow(data))
cortest.bartlett(cor(test_4), n = nrow(data))
```
##KMO test
```{r}
#Train
KMO(train_0)
KMO(train_1)
KMO(train_2)
KMO(train_3)
KMO(train_4)
#Test
KMO(test_0)
KMO(test_1)
KMO(test_2)
KMO(test_3)
KMO(test_4)
```
#Exploratory Factor Analysis (EFA)
```{r}
#0 
fa.parallel(train_0)
#Stu
fa.parallel(train_1)
#2
fa.parallel(train_2)
#3
fa.parallel(train_3)
#4
fa.parallel(train_4)
```
##Fit EFA
```{r}
#0 
efa_0 <- esem_efa(train_0, nfactors = 4)
loadings_0 <- efa_0$loadings
efa_1 <- esem_efa(train_1, nfactors = 4)
loadings_1 <- efa_1$loadings
#2
efa_2 <- esem_efa(train_2, nfactors = 4)
loadings_2 <- efa_2$loadings
#3
efa_3 <- esem_efa(train_3, nfactors = 4)
loadings_3 <- efa_3$loadings
#4
efa_4 <- esem_efa(train_4, nfactors = 4)
loadings_4 <- efa_4$loadings
```


```{r}
loadings_0
loadings_1
loadings_2
loadings_3
loadings_4
```

###Rename columns
```{r}
new_names <- c(log_RT = "(log)Reaction time",
               FIS = "Fluid intelligence score",
               Numeric.memory.Max.digits.remembered.correctly = "Numeric memory: Maximum digits remembered correctly",
               log_TMT.Duration.to.complete.numeric.path = "(log)Trail making test: Duration numeric",
               log_TMT.Duration.to.complete.alphanumeric.path = "(log)Trail making test: Duration alphabetic",
               SDS.Numb.of.symbol.digit.matches.made.correctly = "Symbol digit substitution: Number of correct matches",
               PALNumbofwordpairscorrectlyassociated = "Paired associate learning: Number of correct pairs",
               Tower.rearranging.Numb.of.puzzles.correct = "Tower rearranging: Number of puzzles correct",
               Matrix.PC..Numb.of.puzzles.correctly.solved = "Matrix pattern completion: Number of puzzles correct",
               log1p_Pairs.match.Incorrect.matches..round.2. = "(log1p)Pairs matching: Incorrect matches",
               Picture.vocab.Specific.cognitive.ability = "Picture vocabulary: Specific cognitive ability",
               Prosp.memory.Initial.answer = "Prospective memory: Initial answer")
```
##Create loading tables
```{r}
#0
esem_loadings_0 <- data.table(matrix(round(loadings_0, 4), nrow = 12, ncol = 4))
names(esem_loadings_0) <- c("F1","F2","F3", "F4")
esem_loadings_0$item <- paste0(new_names)
esem_loadings_0 <- melt(esem_loadings_0, "item", variable.name = "latent")
#1
esem_loadings_1 <- data.table(matrix(round(loadings_1, 4), nrow = 12, ncol = 4))
names(esem_loadings_1) <- c("F1","F2","F3", "F4")
esem_loadings_1$item <- paste0(new_names)
esem_loadings_1 <- melt(esem_loadings_1, "item", variable.name = "latent")
#2
esem_loadings_2 <- data.table(matrix(round(loadings_2, 4), nrow = 12, ncol = 4))
names(esem_loadings_2) <- c("F1","F2","F3", "F4")
esem_loadings_2$item <- paste0(new_names)
esem_loadings_2 <- melt(esem_loadings_2, "item", variable.name = "latent")
#3
esem_loadings_3 <- data.table(matrix(round(loadings_3, 4), nrow = 12, ncol = 4))
names(esem_loadings_3) <- c("F1","F2","F3", "F4")
esem_loadings_3$item <- paste0(new_names)
esem_loadings_3 <- melt(esem_loadings_3, "item", variable.name = "latent")
#4
esem_loadings_4 <- data.table(matrix(round(loadings_4, 4), nrow = 12, ncol = 4))
names(esem_loadings_4) <- c("F1","F2","F3", "F4")
esem_loadings_4$item <- paste0(new_names)
esem_loadings_4 <- melt(esem_loadings_4, "item", variable.name = "latent")
```
##Visualize factor loadings
```{r}
#0-4
indices <- 0:4
for (i in indices) {
  # Create a plot name
  plot_name <- paste0("loading_plot_", i)
  # Create a data name
  data_name <- paste0("esem_loadings_", i)
  # Assign the plot to the plot name using assign()
  assign(plot_name, ggplot(get(data_name), aes(x = item, y = value, fill = latent)) + 
           geom_bar(stat = "identity") + coord_flip() + facet_wrap(~ latent, ncol = 1) + 
  theme(legend.position = 'none', panel.grid.major = element_blank(), panel.grid.minor = element_blank(), panel.background = element_rect(fill = "white"),
        axis.title.y = element_text(vjust = 15),
        axis.title.x = element_text(vjust = -0.75, color = "black", size = 20),
        
        plot.title = element_text(color = "black", size = 25),
        plot.subtitle = element_text(color = "black", size = 40),
        
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 8)) + scale_x_discrete(position = "top") +
        # Remove y axis name, change x axis title, and add main title
        ylab("Factor loading") + theme(axis.title.x = element_text(size = 10)) + xlab("") + ggtitle(paste0("Fold ", i)) + theme(plot.title = element_text(hjust = 0.5, size = 15)))
}

loading_plot_0
loading_plot_1
loading_plot_2
loading_plot_3
loading_plot_4
```
##Find referents and build the model
```{r}
#0
referents_0 <- find_referents(efa_0, factor_names = paste0("F",1:4))
model_syntax_0 <- syntax_composer(efa_0, referents_0, only_fix_crossloadings = FALSE)
g_model_0 <- paste(model_syntax_0,
                               "g =~ NA*F1 + F2 + F3 + F4 \n g ~~ 1*g \n", sep = "\n")
#1
referents_1 <- find_referents(efa_1, factor_names = paste0("F",1:4))
model_syntax_1 <- syntax_composer(efa_1, referents_1, only_fix_crossloadings = FALSE)
g_model_1 <- paste(model_syntax_1,
                               "g =~ NA*F1 + F2 + F3 + F4 \n g ~~ 1*g \n", sep = "\n")
#2
referents_2 <- find_referents(efa_2, factor_names = paste0("F",1:4))
model_syntax_2 <- syntax_composer(efa_2, referents_2, only_fix_crossloadings = FALSE)
g_model_2 <- paste(model_syntax_2,
                               "g =~ NA*F1 + F2 + F3 + F4 \n g ~~ 1*g \n", sep = "\n")
#3
referents_3 <- find_referents(efa_3, factor_names = paste0("F",1:4))
model_syntax_3 <- syntax_composer(efa_3, referents_3, only_fix_crossloadings = FALSE)
g_model_3 <- paste(model_syntax_3,
                               "g =~ NA*F1 + F2 + F3 + F4 \n g ~~ 1*g \n", sep = "\n")
#4
referents_4 <- find_referents(efa_4, factor_names = paste0("F",1:4))
model_syntax_4 <- syntax_composer(efa_4, referents_4, only_fix_crossloadings = FALSE)
g_model_4 <- paste(model_syntax_4,
                               "g =~ NA*F1 + F2 + F3 + F4 \n g ~~ 1*g \n", sep = "\n")
```
#Fit the model

- Brown, T. A. (2015). Confirmatory factor analysis for applied research (2nd Edition). New York, NY: Guilford Press.
Hu, L., & Bentler, P. M. (1999). Cutoff criteria for fit indexes in covariance structure analysis: Conventional criteria versus new alternatives. Structural Equation Modeling, 6, 1-55.
- Kline, R. B. (2010). Principles and practice of structural equation modeling (3rd Edition). New York, NY: Guilford Press.
Little, T. D. (2013). Longitudinal structural equation modeling. New York, NY: Guilford Press.

```{r}
#0
model_0_fit <- cfa(g_model_0, std.lv=T, data = train_0)
summary(model_0_fit, fit.measures = T, standardized = T)
#1
model_1_fit <- cfa(g_model_1, std.lv=T, data = train_1)
summary(model_1_fit, fit.measures = T, standardized = T)
#2
model_2_fit <- cfa(g_model_2, std.lv=T, data = train_2)
summary(model_2_fit, fit.measures = T, standardized = T)
#3
model_3_fit <- cfa(g_model_3, std.lv=T, data = train_3)
summary(model_3_fit, fit.measures = T, standardized = T)
#4
model_4_fit <- cfa(g_model_4, std.lv=T, data = train_4)
summary(model_4_fit, fit.measures = T, standardized = T)
```

# Calculate the proportion of variance explained by the g-factor

```{r}
#inspect(model_4_fit, what = "std")$theta - standardized unique (residual) variances of all items 
# 1 - rowSums(...) the variance explained by the g-factor for each item.

total_variance_explained_0 <- 1 - mean(rowSums(inspect(model_0_fit,what="std")$theta))
total_variance_explained_0

total_variance_explained_1 <- 1 - mean(rowSums(inspect(model_1_fit,what="std")$theta))
total_variance_explained_1


total_variance_explained_2 <- 1 - mean(rowSums(inspect(model_2_fit,what="std")$theta))
total_variance_explained_2


total_variance_explained_3 <- 1 - mean(rowSums(inspect(model_3_fit,what="std")$theta))
total_variance_explained_3


total_variance_explained_4 <- 1 - mean(rowSums(inspect(model_4_fit,what="std")$theta))
total_variance_explained_4

mean_total_variance_explained <- mean(c(
  total_variance_explained_0,
  total_variance_explained_1,
  total_variance_explained_2,
  total_variance_explained_3,
  total_variance_explained_4
))
mean_total_variance_explained

```

#lavaan Plot
```{r}
indices <- 0:4
for (i in indices) {
  plot_name <- paste0("lavaanPlot_", i)
  data_name <- paste0("model_", i, "_fit")
  assign(plot_name, lavaanPlot(model = get(data_name), coefs = TRUE,

labels = list(
MatrixPCNumbofpuzzlescorrectlysolved = "Matrix pattern completion",
NumericmemoryMaxdigitsrememberedcorrectly = "Numeric memory",
FIS = "Fluid intelligence score",
ProspmemoryInitialanswer = "Prospective memory",
PALNumbofwordpairscorrectlyassociated = "Paired associate learning",
log_TMTDurationtocompletealphanumericpath = "(log)Trail making test, alphanumeric", SDSNumbofsymboldigitmatchesmadecorrectly = "Symbol digit substitution",
PicturevocabSpecificcognitiveability = "Picture vocabulary",
log_RT = "(log)RT",
log1p_PairsmatchIncorrectmatchesround2 = "(log)Pairs matching",
log_TMTDurationtocompletenumericpath = "(log)Trail making test, numeric",
TowerrearrangingNumbofpuzzlescorrect = "Tower rearranging")
))
}

lavaanPlot_0
lavaanPlot_1
lavaanPlot_2
lavaanPlot_3
lavaanPlot_4
```
#Get g-factor
```{r}
for (i in 0:4) {
  model <- get(paste0('model_', i, '_fit'))
  train_data <- get(paste0("train_", i))
  test_data <- get(paste0("test_", i))
  latents_train <- lavaan::lavPredict(model, newdata = train_data, type = "lv")
  latents_test <- lavaan::lavPredict(model, newdata = test_data, type = "lv")
  
  # Get latents
  assign(paste0("latents_train_", i), latents_train, envir = .GlobalEnv)
  assign(paste0("latents_test_", i), latents_test, envir = .GlobalEnv)
  
  # Extract g
  g_train <- cbind(latents_train[, 5])
  colnames(g_train) <- "g"
  
  g_test <- cbind(latents_test[, 5])
  colnames(g_test) <- "g"
  
  assign(paste0("g_train_", i), g_train, envir = .GlobalEnv)
  assign(paste0("g_test_", i), g_test, envir = .GlobalEnv)
  
  # Save as csv
  #write.csv(g_train, file = file.path(paste0("fold_", i), paste0("g_train_", i, ".csv")), row.names = FALSE)
  #write.csv(g_test, file = file.path(paste0("fold_", i), paste0("g_test_", i, ".csv")), row.names = FALSE)

  #Extract factors
  f_train <- cbind(latents_train[, 1:4])
  f_test <- cbind(latents_test[, 1:4])
  
  assign(paste0("f_train_", i), f_train, envir = .GlobalEnv)
  assign(paste0("f_test_", i), f_test, envir = .GlobalEnv)
  
  # Save as csv
  #write.csv(f_train, file = file.path(paste0("fold_", i), paste0("f_train_", i, ".csv")), row.names=FALSE)
  #write.csv(f_test, file = file.path(paste0("fold_", i), paste0("f_test_", i, ".csv")), row.names=FALSE)
  
}
```